---
title: "Inseason Bonneville Dam Chinook salmon passage predictions"
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: false
    theme: cosmo
params:
  use_dev_version: true
---

```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://privatelands.wdfw.wa.gov/wdfwlogo_clrnotxt.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```

------------------------------------------------------------------------

Last updated `r format(Sys.time(), '%m/%d/%Y')`.

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

if (params$use_dev_version) {
  pkgload::load_all(".")
# } else {
  # library(Inseasonfor)
}




#summer chinook forecast = 38,000, log_sd~ 0.33
#spring chinook forecast = 122500, log_sd~ 0.3

```

```{r data}
#fish data
Bon_cnts<-bon_dat_fun(pred_date=NULL,
                      count_file=NULL,
                      url = "https://www.fpc.org/adults/R_coeadultcount_runsum") |> 
  dplyr::filter(month>=3)

forecastdate <- Bon_cnts |> dplyr::pull(CountDate) |>  max()
forecast_season<-chk_season(forecastdate)
forecast_year<-lubridate::year(forecastdate)
forecast_month<-lubridate::month(forecastdate)
forecast_md<-lubridate::mday(forecastdate)


  season_dates<-data.frame(season=c("spring","summer"),
                           start_m=c("Mar","Jun"),
                           start_md=c(1,16),
                           end_m=c("Jun","Jul"),
                           end_md=c(15,31))
  

bon_yr_seas<-Bon_cnts |> dplyr::filter(season==forecast_season,year==forecast_year)


if(nrow(bon_yr_seas)==0){
  errorCondition("There are no counts for the season of interest")
}

# flow temp data
flow_temp_dat<-data.frame()
num_tries<-0
while((nrow(flow_temp_dat)==0)&num_tries<6){
flow_temp_dat<-get_flow_data(forecastdate = forecastdate,flow_file = NULL) 
num_tries<-num_tries+1

if( num_tries==5 & (nrow(flow_temp_dat)==0) ){
  stop("Unable to get flow temp data from US Army Corps website")
}
}

flow_temp_dat<-flow_temp_dat|> flow_ema_fun()


##add 10-year average to current year flow and temp data for plotting
avg_10yr <- flow_temp_dat |> 
  dplyr::filter(Year < forecast_year & Year >= forecast_year - 10) |> 
  dplyr::group_by(month,md) |> 
  dplyr::summarise(
    `Flow (kcfs)`= mean(cfs_USACoE[Year < forecast_year & Year >= forecast_year - 10], na.rm = TRUE),
     `River temp. (F)`= mean(temp_f_USACE[Year < forecast_year & Year >= forecast_year - 10], na.rm = TRUE),
    .groups="drop") |> 
  dplyr::mutate(type="10yr ave.")

current_yr<- flow_temp_dat |> 
  dplyr::filter(Year == forecast_year ) |> 
  dplyr::select(
    `Flow (kcfs)`=cfs_USACoE,
     `River temp. (F)`= temp_f_USACE,
    flw_date,month,md) |> 
  dplyr::mutate(type="Current yr")


flow_temp_dat2<-
  current_yr |> 
  dplyr::bind_rows(avg_10yr)
  



#ocean cov data
ocean_cov<-ocean_cov_fun(forecast_year) |> 
  dplyr::mutate(Year=Year+2) |> dplyr::select(year=Year,lag2_Spr_NPGO = Spr_NPGO, lag2_Spr_PDO = Spr_PDO, lag2_Sum_NPGO = Sum_NPGO, lag2_Sum_PDO = Sum_PDO)



# #fish data combined with river and ocean covariates for modeling
# fish_river_ocean<-cnts_for_mod_fun(forecastdate,Bon_cnts) |>
#   dplyr::left_join(flow_temp_dat |>
#                      dplyr::filter(
#                        month==forecast_month,
#                                    md==forecast_md) |>
#                      dplyr::select(year=Year,cfs_mean_ema,temp_mean_ema),
#   ) |>
#   dplyr::left_join(
#     ocean_cov
#   )|>
#         dplyr::mutate(
#           cnt_by_flow= cfs_mean_ema*log_cum_cnt,
#           cnt_by_temp= temp_mean_ema *log_cum_cnt,
#         ) |>   dplyr::filter(year<=forecast_year)
# 
# 
# sib_R<-do_sibregresr_fun(fish_river_ocean,cov_vec=c("log_cum_cnt","cnt_by_flow"))
# 
# 
# arima_F<-do_salmonForecasting_fun(data=fish_river_ocean,cov_vec=c("log_cum_cnt","cnt_by_flow"))
# 
#  JF_F<-fit_joint_likelihood(fish_river_ocean ,forecast = 118078,forecast_log_sd = .3)
#  JF_F2<-fit_joint_likelihood2(fish_river_ocean )
# 
# 
#  dplyr::bind_rows(sib_R,arima_F,JF_F,JF_F2) |> View()

```






```{r, echo=FALSE, results='asis'}
cat("# {.tabset}","\n\n")


for (season_i in c("spring","summer")){
  
  seas_end<-(season_dates |> dplyr::filter(season==season_i) |> 
    dplyr::select(end_m, end_md) |> paste(collapse=" ") |> paste(forecast_year,collapse=",") |> lubridate::mdy())
  
  
    seas_start<-(season_dates |> dplyr::filter(season==season_i) |> 
    dplyr::select(start_m, start_md) |> paste(collapse=" ") |> paste(forecast_year,collapse=",") |> lubridate::mdy())
  
  
     #if forecast date greater than the season end date
    if (forecastdate>seas_end){
      forecastdate_i<-seas_end
  ##set date to last day of season
  
    }else{
      
        forecastdate_i<-forecastdate
      }
  
  
  #if forecast date less than the season start date
  ## leave date as is but do not run plotting functions, only blurb.
  
  
 
  

Bon_ch<-Bon_ch_fun(forecastdate_i,Bon_cnts)
Bon_ch_year<-Bon_ch |> dplyr::filter(year==forecast_year)|>
  dplyr::filter(month>=3)




cat("## ",season_i,ifelse(season_i==forecast_season,"{.active}",""),"\n\n")



inital_blurb_fun(Bon_ch|> dplyr::filter(CountDate==forecastdate_i),
                 season_dates=season_dates)


cat("\n\n")

if(forecastdate_i<=forecastdate){

summary_plot_tabs(flow_temp_dat2,Bon_ch,forecastdate_i)


mod_wrapper_fun(forecastdate_i,Bon_cnts,flow_temp_dat,ocean_cov,Bon_ch_year,season_dates=season_dates,season_end_date=seas_end)


bon_ch_tabs(
    Bon_ch_year,
    forecast_year,
    forecastdate_i
)

}
}

cat("# {-}","\n\n")
```


## Data sources

Fish counts are accessed from the [Fish Passage Center](https://www.fpc.org) and river environment data from the [US Army Corps of Engineers Northwest District](https://www.nwd-wc.usace.army.mil/dd/common/dataquery/www/?k=bonneville)

## Additional resources

**Columbia Basin Research** produces spring and fall Chinook run size predictions available [here](https://www.cbr.washington.edu/inseason/adult).

[This site](https://oscrpapps.shinyapps.io/TAC-Inseason-Preds-sh/) shows **spring Chinook run size predictions from ODFW** based on a log-linear regression between end-of-season dam counts and 2 predictors: 1) the total jack dam counts from the previous year, and 2) the cumulative adult counts to date.

[This site](https://oscrpapps.shinyapps.io/tac-inseason-preds-sox/) shows **sockeye run size predictions from ODFW**.


Data from **test fishing** is available [here](https://wdfw.wa.gov/fishing/management/columbia-river/research/test-fishing)

A plethora of information including pre-season **forecasts and fact sheet** with catch updates is available [here](https://wdfw.wa.gov/fishing/management/columbia-river/compact). Note that pre-season forecasts are generally of an index of abundance that includes the Bonneville Dam counts and harvest below Bonneville Dam, so a greater number than what is being predicted here.

**Code** used to generate predictions and render website is available [here](https://github.com/wdfw-fp/Inseasonfor)
