---
title: "Inseason forecasts"
author: "Mark Sorel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

forecastdate <- lubridate::today()-1
forecast_season<-chk_season(forecastdate)
forecast_year<-lubridate::year(forecastdate)
forecast_month<-lubridate::month(forecastdate)
forecast_md<-lubridate::mday(forecastdate)

#summer chinook forecast = 38,000, log_sd~ 0.33
#spring chinook forecast = 122500, log_sd~ 0.3

```

```{r data}
#fish data
Bon_cnts<-bon_dat_fun(pred_date=NULL,
                      count_file="fish_counts.csv",
                      url = "https://www.fpc.org/adults/R_coeadultcount_runsum")


bon_yr_seas<-Bon_cnts |> dplyr::filter(season==forecast_season,year==forecast_year)


if(nrow(bon_yr_seas)==0){
  errorCondition("There are no counts for the season of interest")
}

#max date
counts_through<- bon_yr_seas %>%  dplyr::pull(CountDate) %>% max()

if(counts_through<forecastdate){
  forecastdate<-counts_through
  warning("counts are not available through the provided forecast date. Returning results based on the most updated counts available.")
}

# flow temp data
flow_temp_dat<-get_flow_data(forecastdate = forecastdate,flow_file = "flow_temp_dat.csv") |> 
  flow_ema_fun(start_month = 2)


#ocean cov data
ocean_cov<-ocean_cov_fun(forecast_year) |> 
  dplyr::mutate(Year=Year+2) |> dplyr::select(year=Year,lag2_Spr_NPGO = Spr_NPGO, lag2_Spr_PDO = Spr_PDO, lag2_Sum_NPGO = Sum_NPGO, lag2_Sum_PDO = Sum_PDO)



# #fish data combined with river and ocean covariates for modeling
# fish_river_ocean<-cnts_for_mod_fun(counts_through,Bon_cnts) |>
#   dplyr::left_join(flow_temp_dat |>
#                      dplyr::filter(month==forecast_month,
#                                    md==forecast_md) |>
#                      dplyr::select(year=Year,cfs_mean_ema,temp_mean_ema),
#   ) |>
#   dplyr::left_join(
#     ocean_cov
#   )|>
#         dplyr::mutate(
#           cnt_by_flow= cfs_mean_ema*log_cum_cnt,
#           cnt_by_temp= temp_mean_ema *log_cum_cnt,
#         )
# 
# 
# old_sib_R<-do_sibregresr_fun(fish_river_ocean,cov_vec=c("log_cum_cnt","cfs_mean_ema","cnt_by_temp"))
# 
# 
# new_sib_R<-do_sibregresr_fun(fish_river_ocean,cov_vec=c("log_cum_cnt","cfs_mean_ema","cnt_by_temp"))
# 
# 
# new_arima<-do_salmonForecasting_fun(data=fish_river_ocean,cov_vec=c("log_cum_cnt","cnt_by_flow"))
# 
# old_arima<-do_salmonForecasting_fun(data=fish_river_ocean,cov_vec=c("log_cum_cnt","cfs_mean_ema"=))
```



```{r plot_cnts_env}
flow_temp_dat %>%
  dplyr::select(Date=flw_date ,`Flow (kcfs)`=cfs_mean , `River temp. (F)`=temp_mean) |> 
  dplyr::mutate(Date=as.Date(Date)) |> 
  dplyr::inner_join(
    bon_yr_seas |> 
      dplyr::select(Date=CountDate,`Adult count`=AdultChinook)
  ) |> 
  tidyr::pivot_longer(cols=c(`Adult count`, `Flow (kcfs)`, `River temp. (F)`),names_to="Param",values_to="Value") %>%
  ggplot2::ggplot(ggplot2::aes(x=Date,y=Value,col=Param,shape=Param))+ ggplot2::geom_point(size=2.5,show.legend=FALSE)+ggplot2::geom_line(show.legend=FALSE)+ggplot2::facet_wrap(~Param,ncol=1,scales="free_y")+
  ggplot2::scale_color_manual(values=c("orangered4","navy","olivedrab"))+
  ggplot2::labs(y="",
                col=NULL,
                shape=NULL)+
  ggplot2::theme(legend.key=ggplot2::element_blank())
```

```{r  number_crunching}

Bon_ch<-Bon_ch_fun(forecastdate,Bon_cnts)

Bon_ch_year<-Bon_ch |> dplyr::filter(year==forecast_year)
Bon_ch_day<-Bon_ch_year |> dplyr::filter(CountDate==forecastdate)

highlight_row <- which(Bon_ch_year$CountDate == forecastdate)

```

The cumulative adult Chinook passage through `r forecastdate %>% format("%B %d, %Y")` is `r Bon_ch_day$total` fish.

<!-- The 5-year (`r paste0((forecast_year-5),"--",(forecast_year-1))` ) average proportion of the run that had passed Bonneville through `r counts_through %>% format("%B %d")` is `r round((Bon_ch_day$Ave_5yr)*100,1)`%. Based on the cumulative counts to date and this proportion, the expected run size is `r format(round(Bon_ch_day$pred_Ave_5yr),scientific=FALSE, big.mark=",")` fish. -->


The 10-year (`r paste0((forecast_year-10),"--",(forecast_year-1))` ) average proportion of the run that had passed Bonneville through `r forecastdate %>% format("%B %d")` is `r round((Bon_ch_day$Ave_10yr)*100,1)`%. Based on the cumulative counts to date and this proportion, the expected run size is `r format(round(Bon_ch_day$pred_Ave_10yr),scientific=FALSE, big.mark=",")` fish.


```{r timing_plot, fig.cap= "Percent of the run complete by date, 1995--2024."}
Bon_ch |> dplyr::filter(dplyr::between(year,forecast_year-15,forecast_year-1)) |> dplyr::mutate(date=(as.Date(paste(forecast_year,month,mday,sep="-")))) |>
  dplyr::filter(month>=4) |> 
  ggplot2::ggplot(ggplot2::aes(x=date,y=prop*100))+ggplot2::geom_boxplot(ggplot2::aes(group = date))+ ggplot2::scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b"
  )+ggplot2::geom_vline(ggplot2::aes(xintercept = forecastdate),col="darkred",lty=2,lwd=1.1)+ggplot2::ylab("Percent complete")+ggplot2::theme(axis.title.x = ggplot2::element_blank(),text = ggplot2::element_text(size=18))

```


```{r MAPE_10_year, fig.cap= "Mean absolute percent error of predictions based on 10-year average run timing, by day of year."}
Bon_ch_year |>
  dplyr::filter(month>=4) |> ggplot2::ggplot(ggplot2::aes(y = (MAPE_10yr*100),x=CountDate))+ggplot2::geom_col()+ggplot2::geom_vline(ggplot2::aes(xintercept = forecastdate),col="darkred",lty=2,lwd=1.1)+ggplot2::ylab("Mean absolute percent error (MAPE)")+ggplot2::theme(axis.title.x.bottom = ggplot2::element_blank()) #+ ggplot2::geom_hline(ggplot2::aes(yintercept=30),lty=2)

```


```{r modeling}


model_results<-mod_results(forecastdate = forecastdate,
                           Count_dat = Bon_cnts,
                           River_dat = flow_temp_dat,
                           Ocean_dat = ocean_cov,
                           forecast = ifelse(forecast_season=="spring",122500,38000),
                           forecast_log_sd = ifelse(forecast_season=="spring",0.3,0.33))



# add 10 year timing to model resutls
model_results2<-model_results |> 
  dplyr::bind_rows(
    Bon_ch_year |> 
      dplyr::ungroup()|> dplyr::filter(dplyr::between(CountDate,                                            as.Date(paste0(forecast_year,"-04-05")),
                                               forecastdate)) |>
      dplyr::mutate(`Lo 95`=exp(qnorm(.025,log(pred_Ave_10yr),log_sd_10yr)),
                    `Lo 50`=exp(qnorm(.25,log(pred_Ave_10yr),log_sd_10yr)),
                    `Hi 50`=exp(qnorm(.75,log(pred_Ave_10yr),log_sd_10yr)),
                    `Hi 95`=exp(qnorm(.975,log(pred_Ave_10yr),log_sd_10yr)),
                    mod_type="10-year\nave. timing") |>
      dplyr::select(mod_type,predicted_abundance=pred_Ave_10yr,`Lo 95`:`Hi 95`,MAPE=MAPE_10yr,date=CountDate)
  )




model_results2 |> ggplot2::ggplot(ggplot2::aes(x=date,y=predicted_abundance,color=mod_type))+ggplot2::geom_line()+ggplot2::geom_point()+ggplot2::ylab("Predicted total dam count")+ggplot2::theme(axis.title.x = ggplot2::element_blank())+ ggplot2::scale_y_continuous(labels = scales::unit_format(suffix="k",scale = 1e-3))


model_results2 |> ggplot2::ggplot(ggplot2::aes(x=date,y=predicted_abundance,color=mod_type))+ggplot2::geom_line()+ggplot2::geom_point()+ggplot2::geom_ribbon(ggplot2::aes(ymin=`Lo 95`,ymax=`Hi 95`),alpha=.5,color=NA)+ggplot2::geom_ribbon(ggplot2::aes(ymin=`Lo 50`,ymax=`Hi 50`),alpha=.5,color=NA)+ggplot2::facet_wrap(~mod_type)
```








### 5-year average run timing

#### proportion complete
```{r}
prop5yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,Ave_5yr_10_days_early:Ave_5yr_1_days_early,Ave_5yr,Ave_5yr_1_days_late:Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"5yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)signif(x,2))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  knitr::kable(prop5yr_tab, format = "html") |>
    kableExtra::kable_styling()
)
```

#### prediction
```{r}
pred5yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,pred_Ave_5yr_10_days_early:pred_Ave_5yr_1_days_early,pred_Ave_5yr,pred_Ave_5yr_1_days_late:pred_Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"5yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)format(round(x),scientific=FALSE, big.mark=","))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  knitr::kable(pred5yr_tab, format = "html") |>
    kableExtra::kable_styling()
)
```




# 10-year average run timing


#### proportion complete
```{r}
prop10yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,Ave_10yr_10_days_early:Ave_10yr_1_days_early,Ave_10yr,Ave_10yr_1_days_late:Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"10yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)round(x,2))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  prop10yr_tab |>
    kableExtra::kable_styling()
)
```
#### prediction
```{r}
pred10yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,pred_Ave_10yr_10_days_early:pred_Ave_10yr_1_days_early,pred_Ave_10yr,pred_Ave_10yr_1_days_late:pred_Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"10yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)round(x,2))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  prop10yr_tab |>
    kableExtra::kable_styling()
)
```


