---
title: "Inseason forecasts"
author: "Mark Sorel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

forecastdate <- lubridate::today()-1
forecast_season<-chk_season(forecastdate)
forecast_year<-lubridate::year(forecastdate)
forecast_month<-lubridate::month(forecastdate)
forecast_md<-lubridate::mday(forecastdate)

```

```{r data}
#fish data
Bon_cnts<-bon_dat_fun(pred_date=NULL,
                      count_file="fish_counts.csv",
                      url = "https://www.fpc.org/adults/R_coeadultcount_runsum")





bon_yr_seas<-Bon_cnts |> dplyr::filter(season==forecast_season,year==forecast_year)

if(nrow(bon_yr_seas)==0){
  errorCondition("There are no counts for the season of interest")
}

#max date
counts_through<- bon_yr_seas %>%  dplyr::pull(CountDate) %>% max()


# flow temp data
flow_temp_dat<-get_flow_data(forecastdate = forecastdate,flow_file = "flow_temp_dat.csv") |> 
  flow_ema_fun(start_month = 2)










#fish data combined with river and ocean covariates for modeling
fish_river_ocean<-cnts_for_mod_fun(counts_through,Bon_cnts) |> 
  dplyr::left_join(flow_temp_dat |>
                     dplyr::filter(month==forecast_month,
                                   md==forecast_md) |> 
                     dplyr::select(year=Year,cfs_mean_ema,temp_mean_ema),
  ) |> 
  dplyr::left_join(ocean_cov_fun(forecast_year) |> 
                     dplyr::mutate(Year=Year+2) |> dplyr::select(year=Year,lag2_Spr_NPGO = Spr_NPGO, lag2_Spr_PDO = Spr_PDO, lag2_Sum_NPGO = Sum_NPGO, lag2_Sum_PDO = Sum_PDO)
  )
```


```{r sibregresr}


#' make forecast using sibregresr
#'
#' @param fish_river_ocean data 
#'
#' @return
#' @export
#'
#' @examples
do_sibregresr_fun<-function(data,cov_vec=c("log_cum_cnt","cfs_mean_ema","temp_mean_ema","lag2_Spr_NPGO","lag2_Spr_PDO")){

## data for sibregresr package
sib_reg_dat<-data |> dplyr::mutate(Stock=paste("Bon",forecast_season,"Chk",sep="_")) |> dplyr::select(Stock,ReturnYear=year,Age3=tot_jack ,Age4=tot_adult) |> 
  dplyr::filter(ReturnYear<forecast_year)


sib_reg_cov<-fish_river_ocean |> dplyr::select(ReturnYear=year,all_of(cov_vec))



pen_dlm_forecast_cov<-sibregresr::forecast_fun(
  df = sib_reg_dat,
  include = c("PenDlm"),
  transformation = log,
  inverse_transformation = exp,
  scale_x = TRUE,
  scale_y = TRUE,
  perf_yrs = 15,
  wt_yrs = 1,
  covariates = sib_reg_cov,
  penDLM_formula =formula(paste(c("y ~ x" , cov_vec),collapse=" + "))
)

sibregresr::make_table(pen_dlm_forecast_cov$forecasts,"PenDlm")


forecast<-pen_dlm_forecast_cov$forecasts |> dplyr::filter(Age=="4",ReturnYear==forecast_year,model_name=="PenDlm")

log_pred<-log(forecast$Pred)
log_sd<-forecast$log_sd


  matrix(exp(qnorm(c(.025,.25,.5,.75,.975),log_pred,log_sd)),nrow=1) |>
    `colnames<-`(c("L95","L50","Pred","U50","U95")) 
  
  


info<-pen_dlm_forecast_cov$fits |> dplyr::filter(Age==4,n_years==-1)



coefs<-c(unlist(tail(info$MLE[[1]]$result$obj$report()$coefs,1))) |> `names<-`(c("Int","Lag_log_jack",cov_vec[]))


covs<-c(1,unlist(tail(info$xy_dat[[1]],1)[,c("x",cov_vec)]))|> `names<-`(c("Int","Lag_log_jack",cov_vec[]))

mean_effects<-coefs*covs 

mean_effects |> data.frame() |> t()
}

```



```{r salmonForecast}

# abundance=TOTAL,year=Year,species,period

salmonForecasting_dat<-fish_river_ocean |> dplyr::mutate(species="Bon_Spr",period=1) |> dplyr::select(species,period,year,abundance=tot_adult,log_lag_jack,log_cum_cnt,cfs_mean_ema:lag2_Sum_PDO) |>
  tidyr::fill(c("log_lag_jack",cov_vec[-1]))
dplyr::mutate(
  dplyr::across(c("log_lag_jack",cov_vec[-1]),\(x)c(scale(x)))
)


ARIMA_forecast<-SalmonForecasting::do_forecast(salmonForecasting_dat,
                                               covariates =c("log_lag_jack",cov_vec[-1]),max_vars=3,n_cores=8,do_stacking = FALSE,TY_ensemble=15,write_model_summaries=FALSE,include_mod = TRUE)


best_weighting<-ARIMA_forecast$ens$forecast_skill |> dplyr::filter(grepl("w",model)) |> dplyr::filter(MAPE==min(MAPE)) |> dplyr::pull(model) 

#best pred
pred<-ARIMA_forecast$ens$ensembles |>  dplyr::filter(year==forecast_year,
                                               model== best_weighting) |> 
  dplyr::mutate(month=forecast_month,md=forecast_md,date=forecastdate)


#average coefficients
coef_mat<-ARIMA_forecast$rp$top_mods |> 
  dplyr::mutate(dplyr::across(-mod,unlist)) |> 
  dplyr::filter(year==forecast_year) |> 
  dplyr::pull(mod) |> 
  lapply(\(x)x[[1]]) |> 
  dplyr::bind_rows()

model_weights<-ARIMA_forecast$ens$final_model_weights |> dplyr::pull(substr(best_weighting,1,(nchar(best_weighting)-2)))


ave_coefs<-colSums(coef_mat *
                     model_weights ,na.rm=T)

ave_coefs_non_int_or_ARMA<-ave_coefs[!names(ave_coefs)%in%c("intercept",paste0("ar",1:10),paste0("ma",1:10),paste0("sar",1:10),paste0("sma",1:10))]


covars<-salmonForecasting_dat[salmonForecasting_dat$year==forecast_year,c("log_lag_jack",cov_vec[-1])]


covar_effects<-ave_coefs_non_int_or_ARMA*covars[names(ave_coefs_non_int_or_ARMA)]


ARIMA_forecast$plots_and_tables$Table2
ARIMA_forecast$plots_and_tables$Table3
ARIMA_forecast$plots_and_tables$Table4
ARIMA_forecast$plots_and_tables$Figure1
ARIMA_forecast$plots_and_tables$Figure3



```

```{r}
joint_likelihood_fit<-fit_joint_likelihood(fish_river_ocean,forecast = 122500,forecast_log_sd = 0.3)

joint_likelihood_fit$adrep_est$current_pred |> exp()

paste(round(exp(c(joint_likelihood_fit$adrep_est$current_pred) + c(-1.96,1.96)*c(joint_likelihood_fit$adrep_sd$current_pred))),collapse="--")



```

```{r plot_cnts_env}
flow_temp_dat %>%
  dplyr::select(Date=flw_date ,`Flow (kcfs)`=cfs_mean , `River temp. (F)`=temp_mean) |> 
  dplyr::mutate(Date=as.Date(Date)) |> 
  dplyr::inner_join(
    bon_yr_seas |> 
      dplyr::select(Date=CountDate,`Adult count`=AdultChinook)
  ) |> 
  tidyr::pivot_longer(cols=c(`Adult count`, `Flow (kcfs)`, `River temp. (F)`),names_to="Param",values_to="Value") %>%
  ggplot2::ggplot(ggplot2::aes(x=Date,y=Value,col=Param,shape=Param))+ ggplot2::geom_point(size=2.5,show.legend=FALSE)+ggplot2::geom_line(show.legend=FALSE)+ggplot2::facet_wrap(~Param,ncol=1,scales="free_y")+
  ggplot2::scale_color_manual(values=c("orangered4","navy","olivedrab"))+
  ggplot2::labs(y="",
                col=NULL,
                shape=NULL)+
  ggplot2::theme(legend.key=ggplot2::element_blank())
```

```{r}

Bon_ch<-Bon_ch_fun(forecastdate,Bon_cnts)

Bon_ch_year<-Bon_ch |> dplyr::filter(year==forecast_year)
Bon_ch_day<-Bon_ch_year |> dplyr::filter(CountDate==counts_through)

highlight_row <- which(Bon_ch_year$CountDate == forecastdate)

```

The cumulative adult Chinook passage through `r counts_through %>% format("%B %d, %Y")` is `r Bon_ch_day$total` fish.

The 5-year (`r paste0((forecast_year-5),"--",(forecast_year-1))` ) average proportion of the run that had passed Bonneville through `r counts_through %>% format("%B %d")` is `r round((Bon_ch_day$Ave_5yr)*100,1)`%. Based on the cumulative counts to date and this proportion, the expected run size is `r format(round(Bon_ch_day$pred_Ave_5yr),scientific=FALSE, big.mark=",")` fish.


The 10-year (`r paste0((forecast_year-10),"--",(forecast_year-1))` ) average proportion of the run that had passed Bonneville through `r counts_through %>% format("%B %d")` is `r round((Bon_ch_day$Ave_10yr)*100,1)`%. Based on the cumulative counts to date and this proportion, the expected run size is `r format(round(Bon_ch_day$pred_Ave_10yr),scientific=FALSE, big.mark=",")` fish.


```{r}
Bon_ch_year |> ggplot2::ggplot(ggplot2::aes(y = (MAPE_10yr*100),x=CountDate))+ggplot2::geom_col()+ggplot2::geom_vline(ggplot2::aes(xintercept = counts_through),col="darkred",lty=2,lwd=1.1)+ggplot2::ylab("Mean absolute percent error (MAPE)")+ggplot2::theme(axis.title.x.bottom = ggplot2::element_blank())


Bon_ch |> dplyr::filter(dplyr::between(year,forecast_year-15,forecast_year-1)) |> dplyr::mutate(date=(as.Date(paste(forecast_year,month,mday,sep="-")))) |> ggplot2::ggplot(ggplot2::aes(x=date,y=prop*100))+ggplot2::geom_boxplot(ggplot2::aes(group = date))+ ggplot2::scale_x_date(
  date_breaks = "1 month",
  date_labels = "%b"
)+ggplot2::geom_vline(ggplot2::aes(xintercept = counts_through),col="darkred",lty=2,lwd=1.1)+ggplot2::ylab("Percent complete")

```



### 5-year average run timing

#### proportion complete
```{r}
prop5yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,Ave_5yr_10_days_early:Ave_5yr_1_days_early,Ave_5yr,Ave_5yr_1_days_late:Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"5yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)signif(x,2))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  knitr::kable(prop5yr_tab, format = "html") |>
    kableExtra::kable_styling()
)
```

#### prediction
```{r}
pred5yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,pred_Ave_5yr_10_days_early:pred_Ave_5yr_1_days_early,pred_Ave_5yr,pred_Ave_5yr_1_days_late:pred_Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"5yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)format(round(x),scientific=FALSE, big.mark=","))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  knitr::kable(pred5yr_tab, format = "html") |>
    kableExtra::kable_styling()
)
```




# 10-year average run timing


#### proportion complete
```{r}
prop10yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,Ave_10yr_10_days_early:Ave_10yr_1_days_early,Ave_10yr,Ave_10yr_1_days_late:Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"10yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)round(x,2))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  prop10yr_tab |>
    kableExtra::kable_styling()
)
```
#### prediction
```{r}
pred10yr_tab<-Bon_ch_year |> dplyr::ungroup()|>  dplyr::mutate(Date=format(CountDate,"%d-%b")) |> 
  
  dplyr::select(Date,pred_Ave_10yr_10_days_early:pred_Ave_10yr_1_days_early,pred_Ave_10yr,pred_Ave_10yr_1_days_late:pred_Ave_5yr_10_days_late) |> 
  `colnames<-`(c("Date",paste(10:1,"Days Early"),"10yr Average",paste(1:10,"Days Late"))) |> 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),\(x)round(x,2))) |>  knitr::kable(format = "html") |>
  kableExtra::kable_styling(full_width = TRUE) |>
  kableExtra::column_spec(12, background = "lightgrey")|>
  kableExtra::row_spec(highlight_row, background = "lightgrey")

htmltools::browsable(
  prop10yr_tab |>
    kableExtra::kable_styling()
)
```


